%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%A  matgrp.tex  
%
%
%

\Chapter{matgrp}

This manual describes the {\sf matgrp} package. This package for {\GAP}
provides an interface for Solvable Radical (or Fitting-free) algorithms for
matrix groups, using a recognition tree.

The package also implements routines that make this data structure work,
notably routines for the epimorphism on the radical factor, data structures for subgroups, as well as for polyclic
generating sets for solvable matrix (sub)groups.

The code is still experimental, but should be as stable as the underlying
packages.

It is copyright (C) 2014 by Alexander Hulpke and released under version 2
and version 3 of the GPL. (It is a bad habit to include copies of the GPL in
software, as it can easily be found under
\URL{http://gnu.org/licenses/gpl.html}.)

If you were able to install the prerequisite package {\sf recog}, you will
have no problems in installing {\sf matgrp}, which is plain {\GAP} code.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Usage}

There is basically just one operation of interest. The package implements a
method for `FittingFreeLiftSetup' that applies to matrix groups over finite
fields, or over residue class rings of the integers.

\beginexample
gap> gens:=[ [ [ -28, 3, 9 ], [ 0, -1, -6 ], [ 3, 0, 1 ] ],   
>   [ [ -1, 0, 0 ], [ -9, 1, 3 ], [ -3, 0, -1 ] ], 
>   [ [ 0, 0, 1 ], [ 1, 0, 9 ], [ 0, 1, 0 ] ] ];
gap> g:=Group(List(gens,x->x*One(Integers mod 89))); 
<matrix group with 3 generators>
gap> ffs:=FittingFreeLiftSetup(g);;
gap> Size(Image(ffs.factorhom));
704880
gap> RelativeOrders(ffs.pcgs);
[ 11, 2, 2, 2, 89, 89 ]
gap> h:=Group(List(gens,x->x*One(Integers mod 2403)));
<matrix group with 3 generators>
gap> ffs:=FittingFreeLiftSetup(h);;
gap> Size(Image(ffs.factorhom));
704880
gap> RelativeOrders(ffs.pcgs);
[ 3, 2, 2, 11, 2, 2, 2, 89, 89, 3, 3, 3, 3, 3, 3, 3, 3, 3 ]
\endexample

Once this data structure has been computed, library functionality that uses
it (that is newer versions of Solvable Radical/Fitting free algorithms)
become available. (Note that method selection does not yet catch all these
cases and that in particular `FittingFreeLiftSetup' needs to be called
explicitly at the start to avoid falling back on generic library methods.)

\beginexample
gap> cl:=ConjugacyClasses(g);;
gap> Collected(List(cl,Size));
[ [ 1, 1 ], [ 7920, 1 ], [ 7921, 87 ], [ 704880, 1 ], [ 712890, 87 ], 
  [ 62029440, 1 ], [ 62037272, 3916 ], [ 62734320, 174 ], [ 63447210, 3741 ] ]
gap> RepresentativeAction(g,g.1,g.1^g.2);               
[ [ Z(89)^21, Z(89)^60, 0*Z(89) ], [ Z(89)^22, Z(89)^54, Z(89)^80 ], 
  [ Z(89)^86, Z(89)^34, Z(89)^35 ] ]
gap> g.1^last=g.1^g.2; 
true
gap> m:=MaximalSubgroupClassReps(h);
[ <matrix group of size 38683802512903680 with 20 generators>, 
  <matrix group of size 10550127958064640 with 20 generators>, 
  <matrix group of size 29012851884677760 with 19 generators>, 
  <matrix group of size 14651105610240 with 19 generators>, 
  <matrix group of size 4298200279211520 with 18 generators>, 
  <matrix group of size 58025703769355520 with 20 generators>, 
  <matrix group of size 1289460083763456 with 20 generators>, 
  <matrix group of size 29635190893440 with 20 generators>, 
  <matrix group of size 28976631095808 with 20 generators> ]
gap> s:=HallViaRadical(g,[2])[1];
<matrix group of size 128 with 7 generators>
gap> n:=NormalizerViaRadical(g,s);
<matrix group of size 1408 with 7 generators>
\endexample

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Generic Treatment of Matrix groups}

The ultimate goal of this package surely must be to have {\GAP} work with
matrix groups in the same way as with permutation groups, simply using
recognition as underlying structure. This is not yet feasible at the moment
and will require some larger changes to the system:

\beginitems
-&Currently the `recog' package does not provide a verification routine that
would guarantee a correct recognition tree.

-&There needs to be a way to override the default treatment of matrix groups
using nice monomorphisms, if recognition is available.

-&In matrix groups, an initial (safety) test for element membership is
expensive. If the element is not contained in the group, it even might
require a more robust behaviour of action homomorphisms.
In practice such a test ought to be trivial, as the element is known to be
contained in the group from creation. However matrices in {\GAP} do not
carry ``family'' matrix group with them to make such a membership test easy.
\enditems

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\Section{Error Messages}

In certain situations the current version of the code may produce (one-off)
error messages. Some reasons for this are:

\beginitems
-&The recog package does not yet contain a verify routine.

-&Heuristics for numbers of random elements are not yet optimal, and
  fallback code to treat repeated failure, due to unlucky selection of random
elements, is still missing.

-&There might be errors left in the required packages used for matrix group
  recognition.
\enditems

In general, if a calculation in a `matgrp' routine produces an error, it can
be worth trying the command again, more often than not it will run fine. If
errors arise repeatedly, please inform the package author.

Some error messages issued by the package are:

\beginitems
`Error, Wrong size -- set doall option'&
The selection of random radical elements was not enough to generate the
radical. Set the `doall' option to `true' to enforce use of a proven
generating set.

`no vector found'&
When trying a permutation representation for a constituent of the radical
factor, action on vectors found no suitable candidate.

`some discrepancy happened'&
Recognized group order and computed group order disagree. Likely due to an
error in constructive recognition.

`layerlen'&Computed and acltual size of a solvable layer disagree. Likely
due to an error in the randomized stabilizer chain computation.
\enditems

